Image Forensics using Reflections on Curved Surfaces

### 2014-02-20 paper/cover.tex
added details about camera parameters and counterforensics

\title{Detecting Photo Manipulation using Reflections on Curved Surfaces}
\author{Laura A. Moss}
\date{}
\maketitle

@@

\subsection*{Abstract}
Abstract goes here

>>>>

\title{Image Forensics using Reflections on Curved Surfaces}
\author{Laura A. Moss}
\date{}
\maketitle

@@

\subsection*{Abstract}
Image forensics is the study of finding hidden information in photographs. This emerging field encompasses techniques for determining the source of an image, detecting forgery, and detecting steganography. Because pictures are used as hard evidence in journalism, scientific research, and criminal cases, having a strong arsenal of methods for detecting forgery is crucial. Many such methods rely on geometric properties of the image's content, such as mirrors. This research builds on existing methods for finding inconsistencies in reflections on planar mirrors by exploring techniques that would work for spherical mirrors.

### 2014-02-20 paper/main.tex
added details about camera parameters and counterforensics

\chapter{Prior Work}
\section{Reflections on Planar Surfaces}
\label{sec:planar-reflections}
\begin{figure}
	\centering
    	\includegraphics[width=0.8\textwidth]{planar-vanishing-point-forged}
	\caption{Demonstration of vanishing point inconsistency. The lines corresponding to the man on the left converge to a different vanishing point than those from the rest of the scene, so it is likely that the man and his reflection were edited into the scene. \cite[Fig.~1]{obrien12}}
	\label{planar-vanishing-point-forged}
\end{figure}

\begin{figure}
	\centering
    	\includegraphics[width=0.8\textwidth]{planar-midpoint-genuine}
	\caption{Demonstration of midpoint and vanishing point consistency. The lines converge to a well-defined vanishing point $V$. Additionally, the midpoints in three-dimensional scene space between the real and reflected features plausibly intersect the mirror plane. \cite[Fig.~3]{obrien12}}
	\label{planar-midpoint-genuine}
\end{figure}

\begin{figure}
	\centering
    	\includegraphics[width=0.8\textwidth]{center-of-projection-planar-genuine}
	\caption{Demonstration of center of projection consistency. The pink and blue point clouds represent sets of possible solutions, generated by randomly perturbing the points selected that form the lines. Since the point clouds intersect, there is a possible consistent center of projection, or camera position. \cite[Fig.~6]{obrien12}}
	\label{center-of-projection-planar-genuine}
\end{figure}

@@

\section{Other Geometric Techniques}
Other forensic techniques exist that take a geometric approach to shadow consistency. Instead of determining if there is a consistent vanishing point, or a plausible consistent camera location, these techniques determine if there is a consistent location for a light source. One simple technique for determining shadow consistency, akin to the techniques for reflection consistency, involve drawing rays between points on an object and corresponding points on the shadow. These rays should intersect in a plausible location for the light source (or they should be nearly parallel if the light source is the sun) \cite{farid2010image}. This is dependent on the shadows having easily distinguishable features that can be matched up to the object that is casting the shadow.

\begin{figure}
	\centering
    	\includegraphics[width=0.8\textwidth]{moon-landing-consistent}
	\caption{Based on shadow wedge constraints, this photo of the moon landing has a consistent light source. Original image copyright 1969, NASA. \cite[Fig.~1]{kee2013exposing}}
	\label{moon-landing-consistent}
\end{figure}

@@

\chapter{Mathematical Techniques}
\section{Assumptions}
To simplify the geometry of this analysis, we will restrict the form of our non-planar mirrors to be spherical. Quadric surfaces in general are likely to be good enough for typical use cases because most general curves encountered in the real world can be approximated as either a quadric surface or a set of quadric surfaces.

We also assume the camera used to take any of the images analyzed is a point camera with perspective projection. This is a reasonable assumption considering both the size of most digital camera sensors, and the fact that the camera would have to be far enough away from the subjects of the photograph to capture everything required to perform the analysis.
An important idea to note is that showing that an image is consistent according to these techniques is not sufficient to prove that the image is genuine, because the image could have had some counter-forensic techniques applied or been manipulated in a way that is unrelated to reflective geometry.

\section{Properties of Reflective Geometry}
% TODO expand to sync up nomenclature
Reflection of light on a surface is directly related to the surface normal at the point of reflection. The incident ray (or ray of light coming towards the surface) must make the same angle with the normal as the reflected ray.  Additionally, the incident ray, reflected ray, and normal must all be coplanar, and the reflected and incident rays must be on opposite sides of the normal.

@@

\section{Why not estimate the normal?}
% TODO explain that camera parameters are needed
In general, the types of techniques outlined verify that the incident ray is consistent with the normal at the point of reflection. If the projection of the normal could be estimated on any arbitrary image of a quadric surface, then we could easily have a technique to detect inconsistencies in reflections on most curved surfaces because the projection of the normal and the projection of the incident ray should be colinear. However, when perspective projection is taken into account, it is non-trivial to estimate the normal accurately on an arbitrary visible point on the surface, even for a simple surface like a sphere.

@@

@@

One issue with this naive approach to finding the center is that perspective projection distorts the shape of the sphere, depending on how centered the sphere is in the camera's view. If the sphere is towards the edge of the camera's view, the extent of the visible part of the sphere towards the edge is closer to the camera than the visible extent of the part of the sphere towards the middle of the image. This causes the sphere to look larger towards the edge of the image and therefore the apparent edge of the circle distorts out towards the edge of the image. This effect is described in Figure~\ref{sphere-distortion}. As a result, the outline projected onto the image is slightly elliptical, as in Figure~\ref{sphere-perspective-ortho}.

% TODO IRL example of this phenomenon?
% TODO example with smaller sphere

@@

% TODO measurable error- residuals?

\section{Typical Results for Genuine Images}
% TODO add more figures
\begin{figure}[h]
	\centering
    	\includegraphics[width=0.6\textwidth]{1b-genuine}
	\caption{The chair and the cube are reasonably consistent with the center of the sphere, but the incident rays from the letters appear to intersect at a point slightly offset from the center, despite being unmanipulated.}
	\label{1b-genuine}
\end{figure}

@@

\section{Effects of Forgery}
% address: edited images- shifts forward/back, side to side, farther back, composite,
% why the radius test breaks for certain adjustments

@@

\chapter{Future Work}
The primary flaw in the test for rays going through the center is that there is clearly some offset in the rays, at least for the test set generated by Blender. It is possible that Blender generates the reflections inaccurately, but since the test images were generated using ray tracing (as Blender 2.37+ defaults), it is not clear how or why that would happen. In order to evaluate the validity of the test, more investigation is required into whether or not Blender truly generates images with reflections accurately, and real-life photos with spherical mirrors should be staged as well.

@@

@@

% different quadrics
% real world examples probably require enhancement to be able to distinguish features well enough

@@

%\appendix
%\include{appa}
%\clearpage

>>>>

\chapter{Prior Work}
\section{Similar Geometric Techniques}
There are orensic techniques similar to reflection consistency that take a geometric approach to shadow consistency. Instead of determining if there is a consistent vanishing point, or a plausible consistent camera location (see Section~\ref{sec:planar-reflections}), these techniques determine if there is a consistent location for a light source. One simple technique for determining shadow consistency, akin to the techniques for reflection consistency, involve drawing rays between points on an object and corresponding points on the shadow. These rays should intersect in a plausible location for the light source (or they should be nearly parallel if the light source is the sun) \cite{farid2010image}. This is dependent on the shadows having easily distinguishable features that can be matched up to the object that is casting the shadow.

\begin{figure}[h]
	\centering
    	\includegraphics[width=0.8\textwidth]{moon-landing-consistent}
	\caption{Based on shadow wedge constraints, this photo of the moon landing has a consistent light source. Original image copyright 1969, NASA. \cite[Fig.~1]{kee2013exposing}}
	\label{moon-landing-consistent}
\end{figure}

\section{Reflections on Planar Surfaces}
\label{sec:planar-reflections}
\begin{figure}[h]
	\centering
    	\includegraphics[width=0.8\textwidth]{planar-vanishing-point-forged}
	\caption{Demonstration of vanishing point inconsistency. The lines corresponding to the man on the left converge to a different vanishing point than those from the rest of the scene, so it is likely that the man and his reflection were edited into the scene. \cite[Fig.~1]{obrien12}}
	\label{planar-vanishing-point-forged}
\end{figure}

\begin{figure}[h]
	\centering
    	\includegraphics[width=0.8\textwidth]{planar-midpoint-genuine}
	\caption{Demonstration of midpoint and vanishing point consistency. The lines converge to a well-defined vanishing point $V$. Additionally, the midpoints in three-dimensional scene space between the real and reflected features plausibly intersect the mirror plane. \cite[Fig.~3]{obrien12}}
	\label{planar-midpoint-genuine}
\end{figure}

\begin{figure}[h]
	\centering
    	\includegraphics[width=0.8\textwidth]{center-of-projection-planar-genuine}
	\caption{Demonstration of center of projection consistency. The pink and blue point clouds represent sets of possible solutions, generated by randomly perturbing the points selected that form the lines. Since the point clouds intersect, there is a possible consistent center of projection, or camera position. \cite[Fig.~6]{obrien12}}
	\label{center-of-projection-planar-genuine}
\end{figure}

@@

@@

\chapter{Mathematical Techniques}
\section{Approach}
The goal of this analysis is to find a method for determining reflection consistency that is general enough to not need to know the camera parameters, in the same vein as the existing techniques for planar mirrors. To simplify the geometry of this analysis, we will restrict the form of our non-planar mirrors to be spherical. Quadric surfaces in general are likely to be good enough for typical use cases because most general curves encountered in the real world can be approximated as either a quadric surface or a set of quadric surfaces.

We also assume the camera used to take any of the images analyzed is a point camera with perspective projection. This is a reasonable assumption considering both the size of most digital camera sensors, and the fact that the camera would have to be far enough away from the subjects of the photograph to capture everything required to perform the analysis.

An important idea to note is that showing that an image is consistent according to these techniques is not sufficient to prove that the image is genuine, because the image could have had some counter-forensic techniques applied or been manipulated in a way that is unrelated to reflective geometry.

\section{Properties of Reflective Geometry}
% expanded to sync up nomenclature
Reflection of light on a surface is directly related to the surface normal at the point of reflection. The incident ray (or ray of light coming towards the surface) must make the same angle with the normal as the reflected ray.  Additionally, the incident ray, reflected ray, and normal must all be coplanar, and the reflected and incident rays must be on opposite sides of the normal. In 3D space, we can call the point being reflected $P$ and the reflection point $R$. Then the incident ray in 3D space is $PR$, and the projection of $P$, $R$ and $PR$ onto the image plane are $p$, $r$, and $pr$. $pr$ will be referred to as the projection of the incident ray.

@@

\section{Why not estimate the normal?}
\label{sec:normal-estimate}
In general, the types of techniques outlined verify that the incident ray is consistent with the normal at the point of reflection. If the projection of the normal could be estimated on any arbitrary image of a quadric surface, then we could easily have a technique to detect inconsistencies in reflections on most curved surfaces because the projection of the normal and the projection of the incident ray should be colinear. However, when perspective projection is taken into account, it is non-trivial to estimate the normal accurately on an arbitrary visible point on the surface, even for a simple surface like a sphere.

@@

% explain that camera parameters are needed
If the camera parameters were known (which is outside the scope of this paper), the normal could be estimated anywhere on the visible part of the sphere. Knowing the camera parameters implies that we know exactly where the center of projection is relative to the image plane. If we fit an ellipse to the apparent contour of the sphere (see Section \ref{sec:perspective-distortion} for details on why it is an ellipse), we get an elliptic cone formed by the center of projection and the ellipse \cite[p. 201]{hartley}. There is then a family of spheres that fit into the cone that could possibly have created the apparent contour (with increasing radius as the sphere gets farther away). Since the normals on any sphere in the family map to the image plane the same way, it is sufficient to know what the cone is to find the normals.

\section{Why not find the midpoints?}
In the case of a nonplanar mirror, we lack something that is easily obtainable in the planar case: a vanishing point for the incident rays. Thus, we can't find the midpoints in 3D space for the incident rays for a spherical mirror by the same method. It may be possible to find the midpoint by constraining the 3D incident ray to lie between the rays from the camera to the point and its reflection, then restricting the midpoint to both bisect the incident ray and the ray to be colinear with the normal. Again, however, this requires that the camera parameters be known.

@@

One issue with this naive approach to finding the center is that perspective projection distorts the shape of the sphere, depending on how centered the sphere is in the camera's view. If the sphere is towards the edge of the camera's view, the extent of the visible part of the sphere towards the edge is closer to the camera than the visible extent of the part of the sphere towards the middle of the image. This causes the sphere to look larger towards the edge of the image and therefore the apparent edge of the circle distorts out towards the edge of the image. This effect is described in Figure~\ref{sphere-distortion}. As a result, the outline projected onto the image is slightly elliptical, as in Figure~\ref{sphere-perspective-ortho}. If the camera parameters were known, we could just estimate the normal at any point as outlined in Section~\ref{sec:normal-estimate}.

% IRL example of this phenomenon?
% example with smaller sphere?

@@

% measurable error- residuals?

\section{Typical Results for Genuine Images}
\begin{figure}[h]
	\centering
    	\includegraphics[width=0.6\textwidth]{1b-genuine}
	\caption{The chair and the cube are reasonably consistent with the center of the sphere, but the incident rays from the letters appear to intersect at a point slightly offset from the center, despite being unmanipulated.}
	\label{1b-genuine}
\end{figure}

@@

\section{Effects of Forgery}

@@

\chapter{Future Work}
%TODO counterforensics
The primary flaw in the test for rays going through the center is that there is clearly some offset in the rays, at least for the test set generated by Blender. It is possible that Blender generates the reflections inaccurately, but since the test images were generated using ray tracing (as Blender 2.37+ defaults), it is not clear how or why that would happen. In order to evaluate the validity of the test, more investigation is required into whether or not Blender truly generates images with reflections accurately, and real-life photos with spherical mirrors should be staged as well.

@@

It is easy to envision that a forger could beat these techniques by performing the same tests and adjusting as needed. If they were to splice in an object and its reflection, they could align and scale them such that the projections of the incident rays go through the center of the sphere. This is limited mostly by the fact that in a real world scenario, the forger would be restricted by how the reflective surface in the source image is set up. Nonetheless, more research into counterforensics is recommended.

@@

@@
